---
title: "final_project"
author: "Yu He"
date: "2023-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview of dataset: The Youth Risk Behavior Surveillance System (YRBSS) 2019 high school data for all states

The Youth Risk Behavior Surveillance System (YRBSS) monitors six categories of health-related behaviors that contribute to the leading causes of death and disability among youth and adults, including—

* Behaviors that contribute to unintentional injuries and violence
* Sexual behaviors related to unintended pregnancy and sexually transmitted diseases, including HIV infection
* Alcohol and other drug use
* Tobacco use
* Unhealthy dietary behaviors
* Inadequate physical activity

Database website: https://www.cdc.gov/healthyyouth/data/yrbs/index.htm

Codebook for variable included in the dataset -

* sitename - Name of location covered by a survey
* year - 4-digit year of survey – 1991, 1993, etc
* weight - Analytical weight – used in statistical analyses
* age - Data from:
    How old are you?
    A. 12 years old or younger
    B. 13 years old
    C. 14 years old
    D. 15 years old
    E. 16 years old
    F. 17 years old
    G. 18 years old or older
* sex - Data from:
    What is your sex?
    A. Female
    B. Male
* grade - Data from:
    In what grade are you?
    A. 9th grade
    B. 10th grade
    C. 11th grade
    D. 12th grade
    E. Ungraded or other grade
* race4 - 4-level variable from race and ethnicity questions:
    1 = “White”
    2 = “Black or African American”
    3 = “Hispanic/Latino”
    4 = “All Other Races”
* race7 - 7-level variable from race and ethnicity questions:
    1 = “American Indian/Alaska Native”
    2 = “Asian”
    3 = “Black or African American”
    4 = “Hispanic/Latino”
    5 = “Native Hawaiian/Other Pacific Islander”
    6 = “White”
    7 = “Multiple Races (Non-Hispanic)”
* stheight - Data from:
    How tall are you without your shoes on?
* bmi - Body mass index (BMI)
* bmipct - Student’s BMI percentile
* sexid - Sexual identity:
    1 = "Heterosexual"
    2 = "Gay or Lesbian"
    3 = "Bisexual"
    4 = "Not Sure"
* qnfrcig - Currently smoked cigarettes frequently
* qnothhpl - Used birth control pills; an IUD (such as Mirena or
    ParaGard) or implant (such as Implanon or
    Nexplanon); or a shot (such as Depo-Provera), patch
    (such as OrthoEvra), or birth control ring (such as
    NuvaRing) before last sexual intercourse
* qfoodallergy - Food allergies
* qcurrentasthma - Current asthma
* qncurrentopioid - Currently took prescription pain medicine without a
    doctor's prescription or differently than how a doctor
    told them to use it
* qnmarijuanaschool - Used marijuana on school property

## load libraries
```{r}
library(tidyverse)
library(haven)
library(caret)
library(mice)
```

## data import and process, save the processed sas7dbat data into csv file to improve R runnig time. Only run one time.
```{r}
# set path
setwd("~/Desktop/CU Spring 2023/P8451 intro to machine learning for EPI/p8451_final_project")

# load state data a_m
# df_a_m = read_sas("sadc_2019_state_a_m.sas7bdat") %>% 
#  janitor::clean_names() %>% 
#  select(sitename, year, weight, age, sex, grade, race4, race7, stheight, bmi, bmipct, qnobese, sexid, qnfrcig, 
#         qnothhpl, qfoodallergy, qcurrentasthma, qncurrentopioid, qnmarijuanaschool)

# save the data as a CSV file
# write.csv(df_a_m, "df_a_m.csv", row.names = FALSE)

# load state data n_z
#df_n_z = read_sas("sadc_2019_state_n_z.sas7bdat") %>% 
#  janitor::clean_names() %>% 
#    select(sitename, year, weight, age, sex, grade, race4, race7, stheight, bmi, bmipct, qnobese, sexid, qnfrcig, 
#         qnothhpl, qfoodallergy, qcurrentasthma, qncurrentopioid, qnmarijuanaschool)

# save the data as a CSV file
#write.csv(df_n_z, "df_n_z.csv", row.names = FALSE)
```

## assess the pattern of missing variables
```{r}
# load the processed data
df_part1 = read_csv("df_a_m.csv")
df_part2 = read_csv("df_n_z.csv")

# merge two dataset
df_f_project = bind_rows(df_part1, df_part2)
```


```{r}
# find the number of missing observation in each variables in df_f_project
colSums(is.na(df_f_project))

# find the proportion of missing observation in each variables in df_f_project
colMeans(is.na(df_f_project))
```

```{r}

```

















